---
title: "Introduction to codeminer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{codeminer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(codeminer)
```

# Introduction

The goal of codeminer is to simplify working with clinical codes for research using electronic health records. The workflow is as follows:

1.  Create a local resource containing lookup and mapping tables for various clinical codings systems (e.g. ICD10 and Read codes)
2.  Build clinical code lists for conditions of interest by querying this resource

This vignette demonstrates the above using dummy data included with the package.

Also included are functions for mapping between different clinical coding systems, and using Phecodes[@denny2016; @wu2019] with UK Biobank data. See vignettes `vignette('MAP')` `vignette('caliber')` and `vignette('phecodes')` for further information.

# Build a local clinical codes lookup and mappings resource

The first step is to create a local database containing lookup and mapping tables for various clinical coding systems using `build_database()`.

By default this will download the following resources:

-   UK Biobank resource 592 ([Clinical coding classification systems and maps](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592))

-   UK Biobank [data codings file](https://biobank.ctsu.ox.ac.uk/crystal/exinfo.cgi?src=accessing_data_guide)

-   [Phecode lookup and mapping files](https://phewascatalog.org/) (for ICD9 and ICD10 to phecode)

The tables are imported into R, reformatted, and stored as a named list of data frames:

```{r create_dummy_database}
# Create a temporary database with dummy data
(db_path <- create_dummy_database())
Sys.getenv("CODEMINER_DB_PATH")
```

`codeminer` resolves the database location using the following precedence:

1. The `CODEMINER_DB_PATH` environment variable, if set
2. A default location determined by [`rappdirs::user_data_dir()`](https://rappdirs.r-lib.org/reference/user_data_dir.html)

To persist the database location across sessions, set `CODEMINER_DB_PATH` in your `.Renviron`, e.g. using
[`usethis::edit_r_environ(scope = "project")`](https://usethis.r-lib.org/reference/edit.html?q=edit_r_environ#ref-usage):

```
# ./.Renviron
CODEMINER_DB_PATH=/path/to/codeminer-database.duckdb
```

Alternatively, you can point `codeminer` at a specific database file with `codeminer_connect()`:

```r
codeminer_connect(main = "/path/to/codeminer-database.duckdb")
```

The database is a [duckdb](https://r.duckdb.org/index.html) database. `codeminer` manages the
database connection automatically — you don't need to connect or disconnect manually. You can
check the current connection status with `codeminer_status()`:

```{r codeminer_status}
codeminer_status()
```

# Build a clinical code list

## Explore codes

Codes may be explored with:

- `CODES()`: look up descriptions for a set of code in the given code system type

```{r CODES}
CODES(
  codes = c("E10", "E11"),
  type = "ICD-10"
)
```

- `DESCRIPTION():` search for codes that match a description

```{r DESCRIPTION}
DESCRIPTION(pattern = "cyst", type = "ICD-10")
```

<!-- TODO: add in `CHILDREN()` again, once supported -->

<!-- ```{r CHILDREN} -->
<!-- CHILDREN( -->
<!--   codes = "E10", -->
<!--   type = "ICD-10" -->
<!-- ) -->
<!-- ``` -->
