---
title: "Introduction to codeminer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{codeminer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(codeminer)
```

# Introduction

The goal of codeminer is to simplify working with clinical codes for research using electronic health records. The workflow is as follows:

1.  Create a local resource containing lookup and mapping tables for various clinical codings systems (e.g. ICD10 and Read codes)
2.  Build clinical code lists for conditions of interest by querying this resource

This vignette demonstrates the above using dummy data included with the package.

Also included are functions for mapping between different clinical coding systems, and using Phecodes[@denny2016; @wu2019] with UK Biobank data. See vignettes `vignette('MAP')` `vignette('caliber')` and `vignette('phecodes')` for further information.

# Build a local clinical codes lookup and mappings resource

The first step is to create a local database containing lookup and mapping tables for various clinical coding systems using `build_database()`.

By default this will download the following resources:

-   UK Biobank resource 592 ([Clinical coding classification systems and maps](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592))

-   UK Biobank [data codings file](https://biobank.ctsu.ox.ac.uk/crystal/exinfo.cgi?src=accessing_data_guide)

-   [Phecode lookup and mapping files](https://phewascatalog.org/) (for ICD9 and ICD10 to phecode)

The tables are imported into R, reformatted, and stored as a named list of data frames:

```{r create_dummy_database}
# Create a temporary database with dummy data
(db_path <- create_dummy_database())
Sys.getenv("CODEMINER_DB_PATH")
```

`codeminer` resolves the database location using the following precedence:

1. The `CODEMINER_DB_PATH` environment variable, if set
2. A default location determined by [`rappdirs::user_data_dir()`](https://rappdirs.r-lib.org/reference/user_data_dir.html)

To persist the database location across sessions, set `CODEMINER_DB_PATH` in your `.Renviron`, e.g. using
[`usethis::edit_r_environ(scope = "project")`](https://usethis.r-lib.org/reference/edit.html?q=edit_r_environ#ref-usage):

```
# ./.Renviron
CODEMINER_DB_PATH=/path/to/codeminer-database.duckdb
```

Alternatively, you can point `codeminer` at a specific database file with `codeminer_connect()`:

```r
codeminer_connect(main = "/path/to/codeminer-database.duckdb")
```

The database is a [duckdb](https://r.duckdb.org/index.html) database. `codeminer` manages the
database connection automatically — you don't need to connect or disconnect manually. You can
check the current connection status with `codeminer_status()`:

```{r codeminer_status}
codeminer_status()
```

## Single file vs folder

`CODEMINER_DB_PATH` can point at either a single `.duckdb` file (the default) **or** an existing directory. The directory form holds the per-table data files separately at the folder root, which makes the database easier to inspect, distribute, or sync to remote storage. Choose the layout when you call `build_database()`:

```r
# Single .duckdb file (default)
Sys.setenv(CODEMINER_DB_PATH = "/path/to/codeminer.duckdb")
build_database()

# Folder, data tables as per-table .duckdb files (matches single-file
# query performance; recommended folder layout)
dir.create("/path/to/codeminer-folder")
Sys.setenv(CODEMINER_DB_PATH = "/path/to/codeminer-folder")
build_database(format = "duckdb")

# Folder, data tables as parquet (~35% smaller on disk; slower
# recursive queries like CHILDREN())
dir.create("/path/to/codeminer-parquet")
Sys.setenv(CODEMINER_DB_PATH = "/path/to/codeminer-parquet")
build_database(format = "parquet")
```

`codeminer_connect()` figures out which layout you have automatically — no flags to remember at query time.

# Build a clinical code list

## Explore codes

Codes may be explored with:

- `CODES()`: look up descriptions for a set of code in the given code system type

```{r CODES}
CODES(
  c("E10", "E11"),
  type = "ICD-10"
)
```

- `DESCRIPTION():` search for codes that match a description

```{r DESCRIPTION}
DESCRIPTION(pattern = "cyst", type = "ICD-10")
```

<!-- TODO: add in `CHILDREN()` again, once supported -->

<!-- ```{r CHILDREN} -->
<!-- CHILDREN( -->
<!--   codes = "E10", -->
<!--   type = "ICD-10" -->
<!-- ) -->
<!-- ``` -->

## Reviewing a draft codelist as a tree

A common QA workflow is to draft a broad codelist with parent codes only —
e.g. for diabetes mellitus — and then verify the descendant coverage,
spotting any leaf codes worth including explicitly.

`get_relationship_tree()` returns a flat `list(nodes, edges)` suitable for
downstream hierarchy analysis or tree rendering (e.g. `data.tree`,
`ggraph`, `visNetwork`).

```{r get_relationship_tree}
draft <- c("E10", "E11", "E13", "E14")
tree <- get_relationship_tree(draft, type = "ICD-10")

tree$nodes
tree$edges
```

The `in_input_set` column on `nodes` flags codes that were in the
original draft (`TRUE`) versus codes that were picked up by descendant
expansion (`FALSE`). To inspect the subcodes the expansion added:

```{r get_relationship_tree_expanded}
subset(tree$nodes, !in_input_set)
```

The helper restricts edges to the hierarchical parent/child relationship
type only, and applies `endpoints = "both"` internally — so every edge
endpoint is guaranteed to appear in `nodes` (no dangling edges).

For a defined codelist, the new `codes` argument on `get_lookup_table()`
is the most direct way to pull the matching terms + categories:

```{r get_lookup_table_codes}
get_lookup_table("ICD-10", codes = draft) |> dplyr::collect()
```

# Managing tables

## Adding tables

You can add custom lookup, mapping, and relationship tables to the database with `add_lookup_table()`, `add_mapping_table()`, and `add_relationship_table()`. Each requires a data frame and a metadata object created with the corresponding `*_metadata()` constructor:

```{r add_table}
custom_lookup <- data.frame(
  code = c("CUSTOM1", "CUSTOM2"),
  description = c("Custom code 1", "Custom code 2")
)

add_lookup_table(
  custom_lookup,
  lookup_metadata("custom_codes", lookup_version = "v1")
)

CODES("all", type = "custom_codes")
```

## Removing tables

To remove a table, use the corresponding `remove_*_table()` function with the same identifying keys:

```{r remove_table}
remove_lookup_table("custom_codes", "v1")
```

Removing a table deletes both the data table and its metadata entry. After removal, the same code type and version can be re-added.

## Viewing metadata

Use `get_codeminer_metadata()` to inspect the tables currently in the database:

```{r metadata}
get_codeminer_metadata("lookup")
```

# Version pinning

When multiple versions of a lookup, mapping, or relationship table are available, `codeminer` resolves `"latest"` automatically. The first time a query function resolves `"latest"` for a given code type, the resolved version is cached for the remainder of the session. This avoids repeated informational messages and ensures consistent version usage across a workflow.

You can override this for the current session with `codeminer_set_version()`:

```{r version_pinning}
# Pin lookup and relationship versions for a code type
codeminer_set_version(
  lookup       = c("ICD-10" = "UKB v4"),
  relationship = c("ICD-10" = "UKB v4")
)

# Pin a mapping version (use "from > to" format for the key)
codeminer_set_version(
  mapping = c("Read v3 > ICD-10" = "UKB v4")
)
```

Pins only affect the default `"latest"` resolution. Explicit version arguments always take precedence:

```{r version_pinning_explicit}
# This uses the pinned version for ICD-10:
CODES("E10", type = "ICD-10")

# This ignores the pin and uses "UKB v4" directly:
CODES("E10", type = "ICD-10", lookup_version = "UKB v4")
```

To clear all version selections and return to automatic `"latest"` resolution:

```{r version_pinning_clear}
codeminer_clear_versions()
```

You can also clear versions selectively by code type:

```{r version_pinning_selective, eval = FALSE}
# Clear only the ICD-10 lookup version
codeminer_clear_versions(lookup = "ICD-10")

# Clear lookup and relationship for SNOMED CT
codeminer_clear_versions(
  lookup = "SNOMED CT",
  relationship = "SNOMED CT"
)
```

`codeminer_status()` shows any active versions alongside the connection info.

## Storing version settings

For reproducible analysis, you can store your version pins in a configuration file and load them at the start of a session.

**CSV format** (one row per code type, columns for each table type):

```
code_type,lookup,relationship
ICD-10,UKB v4,UKB v4
Read v3,UKB v4,UKB v4
SNOMED CT,GPS v1,GPS v1
```

```r
cfg <- read.csv("codeminer_versions.csv")
codeminer_set_version(
  lookup       = setNames(cfg$lookup, cfg$code_type),
  relationship = setNames(cfg$relationship, cfg$code_type)
)
```

Mapping pins use a `"from > to"` key format and are best stored in a separate file or in JSON:

**JSON format:**

```json
{
  "lookup": {"ICD-10": "UKB v4", "Read v3": "UKB v4"},
  "relationship": {"ICD-10": "UKB v4"},
  "mapping": {"Read v3 > ICD-10": "UKB v4"}
}
```

```r
cfg <- jsonlite::fromJSON("codeminer_versions.json")
codeminer_set_version(
  lookup       = unlist(cfg$lookup),
  relationship = unlist(cfg$relationship),
  mapping      = unlist(cfg$mapping)
)
```

# Column filters

Some tables contain rows that should be excluded by default — for example, inactive SNOMED CT concepts or approximate code mappings. Column filters (`col_filters`) let table authors declare which columns are filterable, what values are available, and which values should be selected by default. You can override them per call, per session, or per scope; the model in one sentence:

> For each table a query touches: the call's `col_filters` entry, else the session pin, else the metadata defaults — first match wins, replacing the whole set; `NA` means unfiltered at whatever level you apply it.

```r
# Defaults (from metadata / session pins)
CODES("all", type = "SNOMED CT")

# No filtering for this call
CODES("all", type = "SNOMED CT", col_filters = NULL)

# Table-keyed override — reaches every table the query touches,
# e.g. both the mapping table and the target lookup in MAP()
MAP(
  bnf_codes,
  from = "BNF",
  to = "SNOMED CT",
  col_filters = list(
    mapping = list("BNF > SNOMED CT" = list(assured = "Y")),
    lookup  = list("SNOMED CT" = list(moduleId_concept = "999000011000001104"))
  )
)
```

Session pinning works like version pinning (`codeminer_set_col_filters()` / `codeminer_clear_col_filters()`), scoped overrides use `with_col_filters()`, and `get_col_filters()` returns the registered filters in the same shape for amend-and-pass-back workflows. See `vignette("col_filters")` for the full guide.
