---
title: "Column filters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Column filters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(codeminer)

create_dummy_database()
```

Some tables contain rows that should usually be excluded — inactive SNOMED CT
concepts, unassured code mappings, retired descriptions. Column filters
(`col_filters`) let table authors declare which columns are filterable, which
values are available, and which are selected by default; and let you override
those defaults per call, per session, or per scope.

The whole model fits in one sentence:

> **For each table a query touches: the call's `col_filters` entry, else the
> session pin, else the metadata defaults — first match wins, replacing the
> whole set; `NA` means unfiltered at whatever level you apply it.**

# Motivation: filtering every table a query touches

A single query often reads more than one table. `MAP()` reads a mapping table
*and* the target lookup; `CHILDREN()` reads a relationship table *and* the
lookup that describes the results. The `col_filters` argument is table-keyed,
so one argument addresses each table by its own name:

```{r motivation-example}
# `MAP()` reads the Read v3 -> ICD-10 mapping table. Keep only exact + general
# ("E", "G") mappings. The same table-keyed argument can also address the
# *target lookup* by name (demonstrated with the `Demo` lookup below).
MAP(
  "all",
  from = "Read v3",
  to = "ICD-10",
  col_filters = list(
    mapping = list("Read v3 > ICD-10" = list(mapping_status = c("E", "G")))
  )
) |>
  head()
```

The top-level names are the table types (`lookup`, `relationship`,
`mapping`); the keys are code types, or `"from > to"` pairs for mappings;
the leaves are `column = c(values)` selections. Entries for tables the query
does not use are simply ignored, so one filter object can be reused across
different queries.

# Where filters come from

Filters are declared in table metadata, per column, with `values` (all valid
options) and `defaults` (applied when nothing overrides them). A column may
also carry two optional documentation fields, so filters are self-describing:

- `description`: a single string explaining what the column means.
- `value_labels`: a named character vector mapping values to human-readable
  labels (its names must be a subset of `values`; labelling may be partial).

`defaults` is the value set applied by default. To **include every value** by
default, list them all — the column is then unrestricted by default but still
available to narrow. To **exclude everything** by default, use an empty
`defaults` (rarely wanted).

The rest of this vignette uses a small lookup registered on the dummy database.
`status` is filtered to active codes by default; `module` lists both of its
values as `defaults`, so all modules are included by default:

```{r register-demo}
demo_lookup <- data.frame(
  code = c("A", "B", "C"),
  description = c("apple", "banana", "cherry"),
  status = c("1", "1", "0"),
  module = c("core", "ext", "core")
)

add_lookup_table(
  demo_lookup,
  lookup_metadata(
    "Demo",
    col_filters = list(
      status = list(
        values = c("0", "1"),
        defaults = "1",
        description = "Whether the code is active.",
        value_labels = c("1" = "Active", "0" = "Inactive")
      ),
      module = list(
        values = c("core", "ext"),
        defaults = c("core", "ext"), # all values -> all included by default
        description = "Which module the code belongs to."
      )
    )
  )
)
```

Filters on an existing table can be changed without re-adding the data via
`update_lookup_metadata()` (and the mapping/relationship equivalents).

# The `col_filters` argument

Every query function accepts `col_filters`:

- **`"default"`** (the default): session pins where set, else metadata
  defaults.
- **`NULL`** (or `NA`): no filtering for any table this query touches.
- **A table-keyed list** (as above): each entry **replaces** that table's
  pinned/default filters wholesale; unnamed tables keep their pins/defaults.
  `NA` as a table entry un-filters that one table.

```{r arg-examples}
# The dummy Read v3 > ICD-10 mapping declares a mapping_status filter with
# defaults E, G, D. Restrict this call to exact ("E") mappings only:
MAP(
  "all",
  from = "Read v3",
  to = "ICD-10",
  col_filters = list(
    mapping = list("Read v3 > ICD-10" = list(mapping_status = "E"))
  )
) |>
  head()

# No filtering at all for this call:
MAP("all", from = "Read v3", to = "ICD-10", col_filters = NULL) |>
  head()
```

Because an entry replaces the whole set, "tweak one column, keep the rest"
is done in data space: take the current filters, amend, pass back (see
[the round-trip workflow](#amending-filters-the-round-trip-workflow)).

Typos don't fail silently: an entry that matches no registered table, a
column absent from its table, or a value outside a column's registered
`values` each trigger a warning naming what *is* available.

## What a filter value means

The distinction that most often surprises people is **empty selection**
versus **unfiltered** — they are deliberately different. Reading from the
inside out:

**Per column** (inside a table entry):

| You set a column to | Meaning |
|---|---|
| `c("a", "b")` | keep rows where the column is `"a"` or `"b"` (a whitelist) |
| `character(0)` (JSON `[]`) | keep **no** rows — an empty allow-set excludes everything |
| `NA` (JSON `null`) | do **not** filter this column (unfiltered) |
| *column omitted* | not filtered — an overlay only constrains the columns it names |

**Per table** (a key under `lookup` / `mapping` / `relationship`):

| Table entry | Meaning |
|---|---|
| a named list of column filters | **replaces** that table's defaults wholesale |
| `NA` | un-filter the whole table |
| *table omitted* | keeps its pins / metadata defaults |

**Whole argument** (`col_filters =`):

| Value | Meaning |
|---|---|
| `"default"` | session pins where set, else metadata defaults |
| `NULL` / `NA` | no filtering for any table this query touches |
| a table-keyed list | as in the tables above |

So an empty selection is a real, useful state ("exclude everything, return
nothing"), and it is **not** the same as removing the constraint. To stop
filtering a column, set it to `NA` or omit it — not `character(0)`. On the
`Demo` table:

```{r sentinels}
# Default: status filtered to "1"; module lists all values, so all included
CODES("all", type = "Demo")$code

# Empty selection on status excludes everything
CODES(
  "all",
  type = "Demo",
  col_filters = list(lookup = list(Demo = list(status = character(0))))
) |>
  nrow()

# NA leaves status unfiltered (all statuses returned)
CODES(
  "all",
  type = "Demo",
  col_filters = list(lookup = list(Demo = list(status = NA)))
)$code
```

## The round-trip guarantee

Because each column's default is its full applied value set — including columns
that are unrestricted by default (they simply list all their values) —
`get_col_filters()` output passed straight back reproduces the default query
exactly:

```{r roundtrip}
identical(
  CODES("all", type = "Demo")$code,
  CODES("all", type = "Demo", col_filters = get_col_filters())$code
)

# Narrow a column by assigning a subset:
cf <- get_col_filters()
cf$lookup$Demo$module <- "core"
CODES("all", type = "Demo", col_filters = cf)$code
```

Opt-in columns are still fully visible under
`get_col_filters(defaults_only = FALSE)` for discovery — they are only absent
from the applied-defaults shape.

# Session pins and scoped overrides

`codeminer_set_col_filters()` pins filters for the session — the same shape,
the same replacement rule, applied whenever a query runs with
`col_filters = "default"`:

```{r pins}
codeminer_set_col_filters(
  mapping = list("Read v3 > ICD-10" = list(mapping_status = "E"))
)

# The pin now applies to every query that touches this mapping table
nrow(MAP("all", from = "Read v3", to = "ICD-10"))

codeminer_clear_col_filters()
nrow(MAP("all", from = "Read v3", to = "ICD-10"))
```

Because a pin replaces a table's defaults wholesale, the setter tells you
when a pin drops registered default columns. `codeminer_set_col_filters(NA)`
disables all filtering for the session; `codeminer_clear_col_filters()`
returns to the metadata defaults. `codeminer_status()` shows the active
state.

`with_col_filters(col_filters, code)` applies filters for the duration of an
expression, then restores the previous state (even on error):

```{r with}
with_col_filters(
  list(mapping = list("Read v3 > ICD-10" = list(mapping_status = "E"))),
  MAP("all", from = "Read v3", to = "ICD-10")
) |>
  nrow()
```

The scope is the *expression*: every codeminer read made while it evaluates
resolves against the temporary filters, so a multi-step pipeline can be
wrapped as a whole, or step by step:

```r
# One scope for both steps: CHILDREN()'s lookup reads are filtered too
with_col_filters(
  list(lookup = list("SNOMED CT" = list(moduleId_concept = "999..."))),
  {
    mapped <- MAP(bnf_codes, from = "BNF", to = "SNOMED CT")
    CHILDREN(mapped)
  }
)

# Or move the boundary so only MAP() is filtered
with_col_filters(
  list(lookup = list("SNOMED CT" = list(moduleId_concept = "999..."))),
  MAP(bnf_codes, from = "BNF", to = "SNOMED CT")
) |>
  CHILDREN()
```

One subtlety: the scope covers query *construction*. Filters are attached to
a lazy table inside the read call, so a lazy `get_lookup_table()` result
built inside a scope keeps its filters even if you `dplyr::collect()` it
after the scope has ended.

# Amending filters: the round-trip workflow

`get_col_filters()` returns the currently registered default filters for
every table, in exactly the shape the `col_filters` argument (and the pins)
accept. Amend it with plain assignment and pass it back:

```{r round-trip}
cf <- get_col_filters()
print(cf)

# Change one column's selection while keeping everything else
cf$mapping[["Read v3 > ICD-10"]]$mapping_status <- c("E", "G")

nrow(MAP("all", from = "Read v3", to = "ICD-10", col_filters = cf))
```

The same object works with `codeminer_set_col_filters(col_filters = cf)` and
`with_col_filters(cf, ...)`. Deleting a column from the object
(`cf$mapping[["Read v3 > ICD-10"]]$refine_flag <- NULL`) removes that
override, and setting a table's entry to `NA` un-filters the table. Columns
need not have a registered filter specification — any real column of the
table can be filtered (like `moduleId_concept` in the motivating example).

# Where should filters live?

- **Defaults that should always apply** belong in the database metadata
  (`update_*_metadata()`): they travel with the database file, so every R
  process — including fresh parallel workers — gets them on connect.
- **Session-wide tweaks** are pins, set where you connect. Fresh R processes
  start with no pins; the place you connect is the place you pin.
- **Per-call or per-scope deviations** use the `col_filters` argument or
  `with_col_filters()`.

# Discovering available filters

`get_col_filters(defaults_only = FALSE)` returns the full specification —
every filterable column with all its `values` and the `defaults`, plus any
`description` and `value_labels` the filter declares — which is what a UI needs
to render labelled filter choices:

```{r discover}
get_col_filters(defaults_only = FALSE)$mapping[["Read v3 > ICD-10"]]
```
