---
title: "Mapping between clinical coding systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping between clinical coding systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(codeminer)
library(dplyr)

create_dummy_database()
```

# Basic usage

Use `MAP()` to map a vector of codes from one clinical coding system to another:

```{r MAP}
MAP(codes = "X40J4", from = "read3", to = "icd10", map_version = "v0")
```

Use `MAP("all")` to return the entire mapping table:

```{r}
MAP("all", from = "read3", to = "icd10")
```

# Filters

Mapping between different clinical coding systems is not perfect and should be manually reviewed. Some mapping tables
list multiple possible code mappings, with a separate column indicating the type of mapping.[^1] For example, the Read 3
to ICD10 mapping table includes `mapping_status` and `refine_flag` columns:

```{r}
MAP("all", from = "read3", to = "icd10") |> 
  filter(from == "XaIP9")
```

[^1]: Refer to the accompanying documentation for [UKB resource 592](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592)
for further details.

It is important to decide which mappings to include. For example, if no filters are applied then the Read 3 code for sebaceous cyst 'XaIP9' will map to a number of ICD10 codes, some of which are sex-specific ('N508' and 'N948'):

```{r}
MAP("XaIP9", from = "read3", to = "icd10") |> 
  filter(!is.na(GENDER_MASK))
```
